### 2012-08-31 Dail_Mad_ext.tex
formatting

\documentclass{article}

\usepackage[a4paper, hmargin={2.4cm,2.4cm}, vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}

@@

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or management actions. State-space
models, are widely used for these purposes, and they allow for
estimating parameters of classical growth models, such as exponential,
logistic, and Gompertz, while accounting for demographic stochasticity
and observation error. Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet{dail_madsen:2011}. We
expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. Furthermore, we describe classical and
Bayesian methods of parameter estimation. These new developments will
allow researchers to apply these methods to address important
questions regarding the factors affecting spatial and temporal
variation in abundance. We demonstrate these developments by analyzing
data from the North American Breeding Bird Survey. Code to fit these
models using the R package unmarked and JAGS is also provided.
}

@@

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error (de Valpine and Hastings 2002, Buckland et
al. 2004, Dennis et al. 2006). Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow Buckland et
al. (2004) using the definition : ``xxx'' . Typically, the ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let  be
the abundance of a species during year  , and   be the observed data,
which differs from  due to observation error ( ). A simple state space
model that includes exponential growth and process variation ( ) could
be written as
	 	(1)
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions:   and   (de
Valpine and Hastings 2002, Dennis et al. 2006). Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace   with   where A is the area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus . Generally, zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

>>>>

\documentclass[12pt]{article}

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}

@@

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or management actions. State-space
models, are widely used for these purposes, and they allow for
estimating parameters of classical growth models, such as exponential,
logistic, and Gompertz, while accounting for demographic stochasticity
and observation error. Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by
\citet[Biometrics]{dail_madsen:2011}. We
expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. Furthermore, we describe classical and
Bayesian methods of parameter estimation. These new developments will
allow researchers to apply these methods to address important
questions regarding the factors affecting spatial and temporal
variation in abundance. We demonstrate these developments by analyzing
data from the North American Breeding Bird Survey. Code to fit these
models using the R package unmarked and JAGS is also provided.
}

@@

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions:   and   (de
Valpine and Hastings 2002, Dennis et al. 2006). Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace   with   where A is the area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus . Generally, zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.
